A molecular taxonomy of tumors independent of tissue-of-origin

### 2020-08-26 text/manuscript.Rmd

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified twenty-five pathways highly
enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4**). Clusters 3 and 5 were distinguished by metabolic
pathways including RNA and protein biosynthesis (**supplementary figure S4**).
Similarly, cluster 4 was distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (\it{i.e.} too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5**). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others (\it{i.e.} is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin (\it{e.g.} In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.

>>>>

To discover what pathways are most important for clustering, we calculated
percent enrichment for each pathway _within cluster_ relative to all other
clusters combined (see methods) and ranked pathways for each cluster from
highest to lowest enrichment. We visualized the enrichment as a heatmap
(**figure 5A**). Using this approach, we identified twenty-five pathways highly
enriched (enrichment score ≥ 0.3, 95% confidence; see methods) in cluster 2,
nine pathways enriched in cluster 3, twenty-four pathways in cluster 4, ten
pathways in cluster 5, ten pathways in cluster 6, twenty-four pathways in
cluster 7, twenty-four pathways in cluster 8, ten pathways in cluster 9, and no
pathways for clusters 1 and 10 (**supplementary table ST2**). Clusters 7, 8, and
9 had several pathways in common. To explore the specific pathways marking each
cluster, we projected disruptions for each of the 377 pathways onto the UMAP
(**supplementary figure S4** and see interactive media from [**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). 
Clusters 3 and 5 were distinguished by metabolic pathways including RNA and 
protein biosynthesis (**supplementary figure S4**). Similarly, cluster 4 was 
distinguished by mutations affecting regulation of DNA
and histone methylation ("DNA methylation", "PRC2 methylates histones and DNA",
and "Nucleosome assembly"). Clusters 7-9 have in common mutations in
extracellular, intracellular, and immune-related signaling pathways (see
**figure 5B** and **supplementary figure S4**). Cluster 2 had the highest
pathway enrichment levels of the three, having mutations in hedgehog signaling,
"β-catenin degradation", "cellular response to hypoxia", "regulation of cell
cycle" and "apoptosis" among others.

@@

Prior efforts to extract common signatures from pan-cancer datasets met with
difficulty in distinguishing tumor samples from their tissue-specific -omics
data signatures. Given our pathway-disruption based clustering, this raises
the question, are tumor phenotypes driven entirely by common driver genes, or by
“silent” tissue-specific effectors (\it{i.e.} too few samples to detect above
statistical significance thresholds), or a combination of both? To answer this
question, we compared top pathway genes for each cluster relative to the TCGA
background to find differentially mutated genes. We ranked odds ratios and
selected the top ten enriched and depleted genes (pvalue < 0.01) for each
cluster (**Figure 6**; odds ratios plot). Clusters 7 and 8, which shared
multiple enrichment in signaling pathways, are largely driven by mutations in
_PI3K_ and its orthologs and _Ras_ genes, respectively (compare _PIK3CA_ and
_KRAS_ panels of **Supplementary figure S5** and see interactive media from 
[**supplemental file 1**](https://junkdnalab.shinyapps.io/PANCAN_supplemental/)). Interestingly, cluster 9, which
also shared multiple enrichment in signaling pathways with clusters 7 and 8, is
enriched for both _PIK3CA_ and _KRAS_. Clusters 3 and 5, defined by enrichment
in metabolic pathways, had mutations in ribosomal proteins and nuclear pore
complex, respectively. Cluster 4 had mutations in genes responsible for
nucleosome structure. Cluster 2 had mutations in proteasomal subunit genes
involved in protein degradation. We also observed that genes that were enriched
for one cluster are depleted from others (\it{i.e.} is enriched in cluster 6,
but depleted in cluster 7; _PIK3CA_ is enriched in cluster 7, but depleted in
clusters 3 and 8). Next, we investigated the proportion of samples per cancer
type for the significant genes within a cluster (**Figure 6**; heatmap).
Surprisingly, cluster-specific tumors were not predominated by one or more
highly mutated genes across all cancers. Instead, when observing the mutation
rate for these genes within samples that belong to a cluster, the mutation rate
is heterogeneous across tumors by tissue origin (\it{e.g.} In cluster 4, CESC was
enriched for _H2AFX_, OV was enriched for _HIST1H2BD_, and UCEC was enriched for
_HIST1H2AC_). Even among the top most mutationally enriched genes within
clusters there is no global pattern, indicating that our clusters are not driven
by individual genes, but rather networks as a whole. Taken together, our data
identifies a framework of cancer type-specific mutations associated with
specific clusters.
